# Module 3: RNA Expression and Differential Expression

## Lecture
<iframe src="https://docs.google.com/presentation/d/1ufYUvnTMj4Wm8UEbHSdI1JZsfS1ao0uZ/preview" width="640" height="480" allow="autoplay"></iframe>  



## Module 3 - Key concepts
* Differential Gene Expression (DGE) Analysis

## Module 3 - Learning objectives

By the end of this module, you should be able to:

* explain what the goal of DGE analysis is
* understand the steps in a DGE analysis
* be aware of the impact that tool choices can have on the results
* interpret DGE visualizations
* perform a DGE analysis with DESeq2


## Module 3 - Practicle Exercise: Differential Gene Expression (DGE) analysis Lab

First step is to load the required R packages and read the data. For the data, we are using the raw counts table provided at the NCBI.

``` {r setup, message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(DESeq2)

#Indicate where the data is available
url <- "https://www.ncbi.nlm.nih.gov/geo/download/?type=rnaseq_counts&acc=GSE114360&format=file&file=GSE114360_raw_counts_GRCh38.p13_NCBI.tsv.gz"

#Get the data
raw_counts <- fread(url)
# convert the data.table to matrix format
raw_counts = as.matrix(raw_counts)
# check what type of object is the variable raw_counts
class(raw_counts)

# set the gene ID values to be the row names for the matrix
rownames(raw_counts) = raw_counts[, "GeneID"]

# now that the gene IDs are the row names, remove the redundant column that contains them
raw_counts = raw_counts[, colnames(raw_counts) != "GeneID"]

# convert the count values from strings (with spaces) to integers, because originally the gene column contained characters, the entire matrix was set to character
class(raw_counts) = "integer"

# create a simple one column dataframe with the group ID
metaData <- data.frame("Condition" = c("UHR", "UHR", "UHR", "HBR", "HBR", "HBR"))

# convert the "Condition" column to a factor data type
# the alphabetical order of these factors will determine the direction of log2 fold-changes for the genes (i.e. up or down regulated)
metaData$Condition = factor(metaData$Condition, levels = c("HBR", "UHR"))

# set the row names of the metaData dataframe to be the names of our sample replicates from the read counts matrix
rownames(metaData) = colnames(raw_counts)

# check the data
str(raw_counts)
head(raw_counts)
summary(raw_counts)
dim(raw_counts)

# check that names of htseq count columns match the names of the meta data rows
# use the "all" function which tests whether an entire logical vector is TRUE
all(rownames(metaData) == colnames(raw_counts))

```

Look at the R output generated by the previous commands and answer the following questions:

* How many features (genes) are in the raw counts table?
* How many reads were mapped (at most) to each gene in the 50% of the genes with the least expression?
* What's the maximum number of reads mapped to a single gene?
* What's the average number of reads mapped to genes in sample `GSM3140457`?


```{r, message=FALSE, warning=FALSE}
# Now let's create a DESeqDataSet
dds <- DESeqDataSetFromMatrix(raw_counts, colData= metaData, design = ~Condition)

# In the design above we could create more complex experimental designs; for example, if we had a
# two factor experiment where the other factor was sex we could then define  
# design = ~condition+sex. Note that the sex info must be a column in the metaData; otherwise, one 
# gets the error "all variables in design formula must be columns in colData"

# Let's check our dds object
dds

# Let's filter low expressed genes. Since each group has 3 replicates, we will require that genes have at least 10 reads in at least 3 samples. [Remember: this step is not strictly required and 10 is and arbitrary (but reasonable) cut-off value]
smallestGroupSize <- 3
#Here we are counting the columns per row with at least 10 reads. If there are at least 3 columns, the row for that gene will be TRUE and we will keep it
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds <- dds[keep,]

dim(dds)

```
* How many genes were removed because of low expression?

The easiest way to perform DESeq2's default DEG analysis is using the function DESeq, which we are going to use later. This function estimates the scaling factors and the dispersion (both are used for normalizing the data), fits a Negative Binomial Generalized Linear Model (GLM) and performs a Wald statistical test.

However, we are going to explore some of these steps individually so you can see what is happening.


```{r, message=FALSE, warning=FALSE}
# Get the scaling factors
ddsSF <- estimateSizeFactors(dds)
sizeFactors(ddsSF)
# Normalize the data using the scaling factors
normalized_counts <- counts(ddsSF, normalized=TRUE)

#### QUALITY CONTROL
### Transform counts for data visualization
rld <- rlog(ddsSF, blind=TRUE)
plotPCA(rld, intgroup="Condition")

### Extract the rlog matrix from the object
rld_mat <- assay(rld) 
### Compute pairwise correlation values
rld_cor <- cor(rld_mat)

rld_cor

```
Based on the PCA plot, would you expect to find differentially expressed genes (DEGs) between the groups HBR vs UHR? Why?

Based on the correlation values between samples, would you conclude that the assumption about the expression levels of most genes remained the same across groups is correct for these data? Why?

### DGE Analysis

```{r, message=FALSE, warning=FALSE}
## Run DGE analysis with default values
ddsDefault <- DESeq(dds)

resultsNames(ddsDefault)

#Extract the results for the contrast (coefficient) we are interested
resDefault <- results(ddsDefault,c("Condition","HBR","UHR"))
resDefault

# Let's look at the dispersion of the mean of the normalized counts. What we are looking for here
# is that the blue dots are closer to the red line than the black dots, and we expect the 
# dispersion to decrease as the mean of normalized counts increases for each gene.
plotDispEsts(ddsDefault)

#Points will be colored blue if the adjusted p value is less than 0.1.
plotMA(resDefault)

```

Interpret the MA plot. 


```{r, message=FALSE, warning=FALSE}

# Let's try alternative approaches for estimating dispersion
ddsLoc <- DESeq(dds, fitType="local")
resLoc <- results(ddsLoc,c("Condition","HBR","UHR"))
ddsMean <- DESeq(dds, fitType="mean")
resMean <- results(ddsMean,c("Condition","HBR","UHR"))

#Let's visualize the effect of applying different dispersion estimators 
par(mfrow=c(1,2))
xlim <- c(1,1e5); ylim <- c(-1.5,1.5)
plotDispEsts(ddsDefault)
plotMA(resDefault, xlim=xlim, ylim=ylim, main="parametric (default")
par(mfrow=c(1,2))
plotDispEsts(ddsLoc)
plotMA(resLoc, xlim=xlim, ylim=ylim, main="local")
par(mfrow=c(1,2))
plotDispEsts(ddsMean)
plotMA(resMean, xlim=xlim, ylim=ylim, main="mean")


```

* Which dispersion estimator method would you say work best for these data? Why?

### Exercise

Another argument that can be changed in the DESeq function controls how the scaling factors are calculated. The scaling factor calculation method is set with the argument `sfType`. Look at the help for this function (by typing `?DESeq` in the R console) and try one of the alternative approaches, and generate the corresponding MA-plot which should look like one of the plots below. Compare these MA-plots with the MA-plot obtained with the default values. Which method would you expect to deem more genes as differentially expressed?

```{r, message=FALSE, warning=FALSE, echo = FALSE}
 ddsPos <- DESeq(dds, sfType="poscounts")
 resPos <- results(ddsPos,c("Condition","HBR","UHR"))
 plotMA(resPos, xlim=xlim, ylim=ylim, main="poscounts")

 ddsIt <- DESeq(dds, sfType="iterate")
 resIt <- results(ddsIt,c("Condition","HBR","UHR"))
 plotMA(resIt, xlim=xlim, ylim=ylim, main="iterate")
```

### Obtain the DEGs

Let's first see how methodological choices change the results. 

```{r, message=FALSE, warning=FALSE}

# Summary of the results with default values
summary(resDefault)
#How many adjusted p-values were less than 0.1?
sum(resDefault$padj < 0.01, na.rm=TRUE)

# Summary of the results using the mean dispersion estimator approach
summary(resMean)
#How many adjusted p-values were less than 0.1?
sum(resMean$padj < 0.01, na.rm=TRUE)

# Summary of the results using the iterative method to calculate the scaling factors
summary(resIt)
#How many adjusted p-values were less than 0.1?
sum(resIt$padj < 0.01, na.rm=TRUE)

```

* Which method reports a higher number of DEGs? Which method would you say is more conservative?

Let's find out the genes with an adjusted pvalue < 0.01 in common to all three methods.

```{r, message=FALSE, warning=FALSE}

resDefault_Sig <- subset(resDefault, padj < 0.01)
resMean_Sig <- subset(resMean, padj < 0.01)
resIt_Sig <- subset(resIt, padj < 0.01)

GenesInCommon <- intersect(intersect(row.names(resDefault_Sig), row.names(resMean_Sig)),row.names(resIt_Sig))

library(ggVennDiagram)
lst4Venn <- list(Default = row.names(resDefault_Sig), Mean = row.names(resMean_Sig), Iterate = row.names(resIt_Sig))
ggVennDiagram(lst4Venn)


```

* Which genes would you consider the most likely to be truly differentially expressed between the two groups? Why?

Let's create some visualizations and save the results into a csv file.

```{r, message=FALSE, warning=FALSE}
# Let's create a heatmap with the DEGs in the intersection between the three approaches

library(pheatmap)
### Generate a HeatMap  using the metadata data for the sample annotation
pheatmap(rld_mat[GenesInCommon,],
    cluster_rows = T, 
    cluster_cols = F,
    show_rownames = F,
    annotation = metaData, 
    border_color = NA, 
    fontsize = 10
    )

#Plot the counts for the most significant DEG
mostSignificantDEG <- which.min(resDefault$padj)
plotCounts(ddsDefault, gene=mostSignificantDEG, intgroup="Condition")
#Which gene is this one?
resDefault[mostSignificantDEG,]

# What are the raw counts for this gene?
raw_counts[rownames(resDefault[mostSignificantDEG,]),]

#Create a volcano plot
threshold_DE <- resDefault$padj < 0.01

ggplot(resDefault) +
    geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = threshold_DE)) +
    ggtitle("Volcano Plot") +
    xlab("log2 fold change") + 
    ylab("-log10 adjusted p-value") +
    theme(legend.position = "none")

#Order the genes by the adjusted pvalue
resOrdered <- resDefault[order(resDefault$padj),]

resOrdered[1:10,]

#What are the columns?
mcols(resOrdered)$description

#Let's save the complete results
write.csv(as.data.frame(resOrdered),  file="DGE_analysis_results.csv")

```
* Can you locate in the volcano plot the gene with the lowest adjusted p-value?
* How many of the top 10 most significant genes are down in the HBR condition?


### Exercise

Save only the results for the genes in the intersection of the three methods in a csv file named "GenesInCommon_results.csv".

## DESeq2 resources

The [DESeq2 bioconductor vignette](https://bioconductor.org/packages//release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html) is a great resource to dive deeper into DESeq2's functionality.
