# Module 4: Pseudo-alignment and enrichment analysis

## Lecture

<iframe src="https://docs.google.com/presentation/d/1A6x8O0ae_uh1w9_zeNX3pyom4OJqcTR8/preview" width="640" height="480" allow="autoplay"></iframe>  

## Lab


In this tutorial you will perform functional enrichment analysis using:

1. gProfiler
2. clusterProfiler

**Required packages**

```{r setup, message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(gprofiler2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(dplyr)
```


```{r , message=FALSE, warning=FALSE,echo=FALSE}
library(DESeq2)
url <- "https://www.ncbi.nlm.nih.gov/geo/download/?type=rnaseq_counts&acc=GSE114360&format=file&file=GSE114360_raw_counts_GRCh38.p13_NCBI.tsv.gz"

raw_counts <- fread(url)

# convert the data.table to matrix format
raw_counts = as.matrix(raw_counts)
class(raw_counts)

# set the gene ID values to be the row names for the matrix
rownames(raw_counts) = raw_counts[, "GeneID"]

# now that the gene IDs are the row names, remove the redundant column that contains them
raw_counts = raw_counts[, colnames(raw_counts) != "GeneID"]

# convert the count values from strings (with spaces) to integers, because originally the gene column contained characters, the entire matrix was set to character
class(raw_counts) = "integer"

# view the first few lines of the gene count matrix
#head(raw_counts)

# create a simple one column dataframe to start
metaData <- data.frame("Condition" = c("UHR", "UHR", "UHR", "HBR", "HBR", "HBR"))

# convert the "Condition" column to a factor data type
# the arbitrary order of these factors will determine the direction of log2 fold-changes for the genes (i.e. up or down regulated)
metaData$Condition = factor(metaData$Condition, levels = c("HBR", "UHR"))

# set the row names of the metaData dataframe to be the names of our sample replicates from the read counts matrix
rownames(metaData) = colnames(raw_counts)

# view the metadata dataframe
#head(metaData)

# check that names of htseq count columns match the names of the meta data rows
# use the "all" function which tests whether an entire logical vector is TRUE
all(rownames(metaData) == colnames(raw_counts))

dds = DESeqDataSetFromMatrix(countData = raw_counts, colData = metaData, design = ~Condition)

# run the DESeq2 analysis on the "dds" object
dds = DESeq(dds)

# view the first 5 lines of the DE results
res = results(dds)

#define DEGs
deg=subset(res,res$padj<0.01)

```


**Enrichment analysis with gProfiler**

Documentation available : [gProfiler](https://biit.cs.ut.ee/gprofiler/gost)

```{r , message=FALSE, warning=FALSE}
gost.res = gost(rownames(deg),
                 organism = "hsapiens",
                 significant = TRUE,
                 measure_underrepresentation=F,
                 user_threshold = 0.05,
                 correction_method = "fdr",
                 evcodes = T)
```


Manhattan plot for all the results

```{r , message=FALSE, warning=FALSE}

gostres = gost.res$result
# filter results to reduce to less specialized or general terms with at least 2 genes from our list
gostres_filt = gostres[gostres$term_size > 15 & gostres$term_size < 2000 & gostres$intersection_size > 2,]

#trasnform score with log10, higher is better
gostres_filt$log10=-1 * log10(gostres_filt$p_value)

#order results by decreasing score value
gostres_filt=gostres_filt[order(gostres_filt$log10,decreasing=F),]

#Manhattan plot
gostplot(gost.res)

```




**Barplot results for each database**

```{r , message=FALSE, warning=FALSE}


# barplots for top 10 KEGG pathways
res_table = gostres_filt[gostres_filt$source=="KEGG",][1:10, ]
 
#modify margins for long terms name
par(mar = c(5, 20, 4, 2))

barplot(
  res_table$log10,
  horiz = TRUE,
  names.arg = res_table$term_name,
  col = "steelblue",
  xlab = "-log10(padj)",
  xlim = c(0, ceiling(max(res_table$log10))),
  main = "Top 10 enriched terms",
  las = 1,      # horizontal labels
  cex.names = 0.8  # text size
)

```


**Gene Ontology Enrichment with clusterProfiler**

clusterProfiler requires Entrez Gene IDs.


- GO Biological Process Enrichment

```{r}
ego_bp = enrichGO(
  gene          = rownames(deg),
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

head(as.data.frame(ego_bp))
```

- Visualize GO Results

```{r fig.width=10, fig.height=6}
barplot(
  ego_bp,
  showCategory = 15
)
```

```{r fig.width=10, fig.height=8}
dotplot(
  ego_bp,
  showCategory = 20
)
```

- KEGG Pathway Enrichment

```{r}
ekegg = enrichKEGG(
  gene = rownames(deg),
  organism = "hsa",
  pvalueCutoff = 0.05
)

head(as.data.frame(ekegg))
```

- Visualize KEGG Results

```{r fig.width=10, fig.height=6}
barplot(
  ekegg,
  showCategory = 15
)
```

```{r fig.width=10, fig.height=8}
dotplot(
  ekegg,
  showCategory = 20
)
```


